---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rstandev

<!-- badges: start -->
[![R-CMD-check](https://github.com/andrjohns/rstan.dev/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/andrjohns/rstan.dev/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

An experimental, header-only interface to Stan in R

## Usage

Initialise and compile a new stanmodel object:

```{r}
library(rstandev)

model = "
functions {
  row_vector rtn_rowvec(row_vector x) {
    return x;
  }
}

data {
  int<lower=0> N;
  array[N] int<lower=0,upper=1> y; // or int<lower=0,upper=1> y[N];
}
parameters {
  real<lower=0,upper=1> theta;
}
model {
  theta ~ beta(1,1);  // uniform prior on interval 0,1
  y ~ bernoulli(theta);
}

generated quantities {
  vector[N] log_lik;
  for (n in 1:N) {
    log_lik[n] = bernoulli_lpmf(y[n] | theta);
  }
}
"

stmod <- stanmodel$new(model_code = model)
stmod$compile()
```

The compiled `stanmodel` object can then call `log_prob` and `grad_log_prob` as long as data is also passed:
```{r}
stmod$log_prob(list(N=2, y=0:1), upars = 0.1)
stmod$grad_log_prob(list(N=2, y=0:1), upars = 0.1)
```

As well as `constrain_` and `unconstrain_pars`:
```{r}
stmod$constrain_pars(list(N=2, y=0:1), upars = 2.5)
stmod$unconstrain_pars(list(N=2, y=0:1), list(theta = 0.1))
```

Any model functions can be exposed using the `expose_functions()` member. By default, these functions will be added as members to the source `stanmodel` object, but they can be exported to the global environment using the `global = TRUE` flag:

```{r}
stmod$expose_functions()
stmod$rtn_rowvec(c(1, 2, 3))
```

Sampling can then be performed using the `sample()` member:
```{r}
samp <- stmod$sample(
  data = list(N=2, y=0:1),
  num_chains = 4,
  num_threads = 8, refresh = 1000)
```
Similarly for `optimize()` and `variational()` members.

The generated `stanfit` object can then be summarised with the `summary()` function:
```{r}
samp$summary()
```

And the ELPD-LOO calculated with the `loo()` member:
```{r}
samp$loo()
```

## Installation

You can install the development version of rstandev from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("andrjohns/sundialr")
devtools::install_github("andrjohns/rstandev")
```
